---
title: "Allele-specific analysis of RNA-seq data using technical replicates"
output:
  html_document:
    df_print: paged
---

We start by loading libraries and functions we will need:

```{r, warning=FALSE, warning=FALSE}
library("tidyverse")
library("knitr")
source("yet_another_functions_up_to_date_24112018.R")
source("PerformDiffAIAnalysisFor2Conditions.R")
```

##5aza data

To demonstrate how the pipeline works, we are going to use data from RNA-seq experiment aimed to study effect of 5aza treatment on expression and allele-speccific expression. Let's look at the data first:

###Experiment design

5 samples of 4.11 Abelson clones:

name     | description  | number of repicates
------------- | -------------
control  | Untreated Day 0  | 2
DMSO  | DMSO control Day 2  | 2
low  | low 5aza treatment of 0.2 µM  | 2
medium | medium 5aza treatment of 2 µM  | 5
high  | high 5aza treatment of 10 µM  | 2

The first step is to create design matrix describing the experiment:

```{r, message=FALSE, warning=FALSE}
replicateNumbers <- c(2,2,2,5,2)
replicateNames <- c("control","DMSO","low","medium","high")
rowsSp <- data.frame(matrix(lapply(1:length(replicateNumbers), function(x){(2*sum(replicateNumbers[1:x-1])+1):(2*sum(replicateNumbers[1:x]))}), nrow=length(replicateNumbers), byrow=T),stringsAsFactors=FALSE)
colnames(rowsSp) <- "replicateCols"
colExp <- data.frame(matrix(lapply(1:length(replicateNumbers), function(x){(sum(replicateNumbers[1:x])-replicateNumbers[x]+1):(sum(replicateNumbers[1:x]))}), nrow=length(replicateNumbers), byrow=T),stringsAsFactors=FALSE)
colnames(colExp) <- "replicateNums"
designMatrix <- cbind(replicateNames, replicateNumbers, rowsSp, colExp)
kable(designMatrix)
```

Now we can look at the data:

```{r, message=FALSE, warning=FALSE}
inTabs <- "~/Dropbox (Partners HealthCare)/replicates_ASE/data/5aza/pr_20180714_ISEKI_processed_gene_extended2.txt"
geneCountTab <- GetGatkPipelineTabs(inTabs, replicateNumbers, multiple = F)
head(geneCountTab)
```

We can calculate allelic imbalances for all genes for all experimental conditions pooling technical replicates:
AI = \frac{maternal counts}{maternal counts + paternal counts}

```{r, message=FALSE, warning=FALSE}
thr <- 10
geneCountTab_10 <- geneCountTab
nameColumns <- function(exp_n, rep_n)  {
  paste0("exp", rep(exp_n, 2*rep_n), "_rep", rep(1:rep_n, each = 2), "_", rep(c("ref", "alt"), rep_n))
}
colnames(geneCountTab_10)[2:27] <- c(nameColumns(1,2), nameColumns(2,2), nameColumns(3,2), nameColumns(4,5), nameColumns(5,2))
geneCountTab_10 <- geneCountTab_10 %>% dplyr::filter(exp1_rep1_ref + exp1_rep1_alt >=thr & exp2_rep1_ref + exp2_rep1_alt >=thr & 
                             exp3_rep1_ref + exp3_rep1_alt >=thr & exp4_rep1_ref + exp4_rep1_alt >=thr & 
                             exp5_rep1_ref + exp5_rep1_alt >=thr & exp1_rep2_ref + exp1_rep2_alt >=thr & 
                             exp2_rep2_ref + exp2_rep2_alt >=thr & exp3_rep2_ref + exp3_rep2_alt >=thr & 
                             exp4_rep2_ref + exp4_rep2_alt >=thr & exp5_rep2_ref + exp5_rep2_alt >=thr)
aiTable <- do.call(cbind, lapply(1:length(replicateNumbers), function(x){
      round(CountsToAI(geneCountTab_10, reps = unlist(designMatrix$replicateNums[x])),3)
    }))
aiTable <- cbind(geneCountTab_10$ensembl_gene_id, aiTable)
colnames(aiTable) <- c("ensembl_gene_id", designMatrix$replicateNames)
kable(head(aiTable))
write.table(aiTable, file = "~/Dropbox (Partners HealthCare)/replicates_ASE/data/5aza/ai_full.txt", row.names = F, quote = F, sep="\t")
```


Then get row-concatenated table of pairewise delta AI for all the 3 experiments:

```{r, warning=FALSE, message=FALSE}
dfDeltaAIPairwise <- do.call(rbind, lapply(1:length(replicateNumbers), function(x){
      CreateMergedDeltaAIPairwiseDF(geneCountTab, mlns=F, repnums = unlist(designMatrix$replicateNums[x]), what=designMatrix$replicateNames[x], thr=10)
    }))
head(dfDeltaAIPairwise)
```



```{r}
P = pnorm(seq(0.4,2,0.2))
P = P - (1-P) 
DF = dfDeltaAIPairwise
DF$geneORsnp = "Gene" 
DF$group = paste(DF$what, DF$geneORsnp, DF$ij)
head(DF)

DFQ = do.call(rbind, lapply(unique(DF$group), 
                            function(r){
                              CreateObservedQuartilesDF(DF[DF$group == r, ], 
                                                        P, ep=1.3, logbase=T, 
                                                        coverageLimit=2000, group=r)
                            })
) 
head(DFQ)
DFQ$ij = sapply(as.character(DFQ$group), function(x){paste(unlist(strsplit(x, ' '))[3:5], collapse = ' ')})
```


```{r}
groupsIntercepts = lapply(unique(DFQ$Q), function(q){
  df = DFQ[DFQ$Q == q, ]
  res = sapply(unique(df$ij), function(x){ # проход по всем парам
      FitLmIntercept(df[df$ij == x, ], 40, morethan=10, logoutput=T)
    })
  names(res) = unique(df$ij)
  2**res # вернулись от логарифма обратно
})

coeffDF <- data.frame(t(do.call(rbind, groupsIntercepts)))
colnames(coeffDF) <- paste0("q_", round(as.numeric(as.character(unique(DFQ$Q))),3))
```

Add names column:

```{r}
coeffDF$exp <- rep(designMatrix$replicateNames, choose(designMatrix$replicateNumbers, 2))
coeffDF <- coeffDF[,c(dim(coeffDF)[2],1:(dim(coeffDF)[2]-1))]
```


Let's compare conditions DMSO (experiment 2) and medium 5aza treatment (experiment 4). We will construct 95%-CIs around AIs and get the resulting classification into genes demonstrating difference in AI between conditions (based on non-overlapping CIs):

```{r}
DiffAI_DMSO_medium <- PerformDiffAIAnalysisFor2Conditions(geneCountTab, unlist(designMatrix$replicateNums[2]), unlist(designMatrix$replicateNums[4]), Q=0.95)
head(DiffAI_DMSO_medium[, c(1,2,4,10)])
print(paste0("number of genes with differential AI ", table(DiffAI_DMSO_medium$diffAI)[2]))
print(paste0("number of genes with no differential AI ", table(DiffAI_DMSO_medium$diffAI)[1]))
```

