---
title: "Allele-specific analysis of RNA-seq data using technical replicates"
output:
  rmarkdown::github_document
---

We start by loading the libraries and functions we will need:

```{r include = FALSE}
library("tidyverse")
library("knitr")
source("ASE_functions.R")
source("PerformAIAnalysis.R")
```

##5aza data

To demonstrate how the pipeline works, we are going to use data from RNA-seq experiment aimed to study effect of 5aza treatment on expression and allele-specific expression. Let's look at the data first:

###Experiment design

5 samples of 4.11 Abelson clones:

name     | description  | number of repicates
------------- | ------------- | -------------
control  | Untreated Day 0  | 2
DMSO  | DMSO control Day 2  | 2
low  | low 5aza treatment of 0.2 µM  | 2
medium | medium 5aza treatment of 2 µM  | 5
high  | high 5aza treatment of 10 µM  | 2

The first step is to create design matrix describing the experiment:

```{r, message=FALSE, warning=FALSE}
experimentNames <- c("control","DMSO","low","medium","high")
techReps <- c(2,2,2,5,2)
designMatrix <- BuildDesign(experimentNames, techReps)
```

###Load the data

Now we can look at the data. Here is the *geneCountTab* dataframe with counts for all genes, all replicates and conditions:

```{r, message=FALSE, warning=FALSE}
inTabs <- "../../../data/5aza/pr_20180714_ISEKI_processed_gene_extended2.txt"
geneCountTab <- GetGatkPipelineTabs(inTabs, designMatrix$techReps, multiple = F)
head(geneCountTab)
```

###Look at AI correlations

We can calculate allelic imbalances for all genes for all experimental conditions pooling technical replicates:
$$AI=\frac{maternal\ counts}{maternal\ counts + paternal\ counts}$$

```{r, message=FALSE, warning=FALSE}
aiTable <- do.call(cbind, lapply(1:length(designMatrix$techReps), function(x){
      round(CountsToAI(geneCountTab, reps = unlist(designMatrix$replicateNums[x]),thr=10),3)
    }))
aiTable <- data.frame(geneCountTab$ensembl_gene_id, aiTable)
colnames(aiTable) <- c("ensembl_gene_id", designMatrix$experimentNames)
head(aiTable)
```

Let's visualize AI correlations, for example between AI estimates for DMSO (experiment 2) and high 5aza treatement(experiment 5):

```{r, message=FALSE, warning=FALSE}
ggplot(aiTable, aes_string(x=as.name(designMatrix$experimentNames[2]), y=as.name(designMatrix$experimentNames[5]))) +
  geom_point(size=0.5) +
  theme_bw() +
  coord_fixed() 
```



###Differential AI analysis

Let's compare conditions DMSO (experiment 2) and high 5aza treatment (experiment 5). We will construct 95%-CIs around AIs and get the resulting classification finding the genes demonstrating difference in AI between conditions (based on non-overlapping CIs).

First we get 95% CIs:

```{r, message=FALSE, warning=FALSE}
CIs_DMSO <- PerformCIAIAnalysis(geneCountTab, unlist(designMatrix$replicateNums[2]), Q=0.95, EPS=1.3, thr=NA, fullOUT=F)
CIs_high <- PerformCIAIAnalysis(geneCountTab, unlist(designMatrix$replicateNums[5]), Q=0.95, EPS=1.3, thr=NA, fullOUT=F)
CIs <- merge(CIs_DMSO[,c("ID", "meanAI", "pm")], CIs_high[,c("ID", "meanAI", "pm")], by="ID")
colnames(CIs) <- c("ID", "meanAI_DMSO", "pm_DMSO", "meanAI_high", "pm_high")
head(CIs)
```

Now let's run the differential analysis, using coverage threshold=40 and Bonferroni p-value correction:

```{r, message=FALSE, warning=FALSE}
thr=40
DiffAI_DMSO_high <- PerformDiffAIAnalysisFor2Conditions(geneCountTab, 
                                                          vect1CondReps = unlist(designMatrix$replicateNums[2]), 
                                                          vect2CondReps = unlist(designMatrix$replicateNums[5]), 
                                                          Q=0.95,
                                                          thr=thr, 
                                                        EPS=1.3, 
                                                        fullOUT=F
                                                        )
head(DiffAI_DMSO_high[,c("ID","diffAI")])
```

We will define a gene as demonstrating differential allelic imbalance between two conditions if this gene has non-intersecting CIs of allelic imbalance estimates in these conditions and if the difference between these allelic imbalance estimates is bigger than some cutoff (set to 0.2 here). The cutoff value is set arbitary depending on the biological question we are asking.

```{r, message=FALSE, warning=FALSE}
minDifference <- 0.2
GenesAIDiff <- DiffAI_DMSO_high[!is.na(DiffAI_DMSO_high$diffAI), ]
GenesAIDiff$deltaAI <- abs(GenesAIDiff$meanAI1 - GenesAIDiff$meanAI2)
GenesAIDiff <- GenesAIDiff[GenesAIDiff$diffAI==T & GenesAIDiff$deltaAI>=minDifference, ]
GenesAIDiff <- GenesAIDiff[order(GenesAIDiff$deltaAI, decreasing = T),]
head(GenesAIDiff[, c(1,4,10)])
print(paste0("Number of genes demonstrating differential allelic imbalance: ", length(GenesAIDiff$ID)))
```
